---
title: "Intro to R workshop"
author: "Jeremy Buhler and Albina Gibadullina"
date: "November 26, 2020"
output: html_document
---

# Intro to R and RStudio

## Preparing the data

Install R packages
```{r}
#install.packages("dplyr")
#install.packages("gglot2")
#install.packages("gapminder")
```


Load R packages
```{r}
library(dplyr)
library(ggplot2)
library(gapminder)
```


Check the description of the "gapminder" dataset
```{r}
#?gapminder
```


Import data from the built-in "gapminder" dataset
```{r}
countries <- gapminder # create a data frame
head(countries) #see the first six rows of the dataframe 
```


See the entire dataset
```{r}
#View(countries)
```


Check the data structure
```{r}
str(countries)
```


Summarize the dataframe
```{r}
summary(countries)
```

Check if the dataframe is complete:

Count the number of observations for each available year
```{r}
table(countries$year)
```


Count the number of observations in each continent for each available year 
```{r}
table(countries$continent, countries$year)
```




## Data Manipulation with dplyr


### Select function

Select specific columns such as country, year, lifeExp
```{r}
select(countries, country, year, lifeExp)
```


Select all columns except for continent and gdpPercap
```{r}
select(countries, -continent, -gdpPercap) # use subtraction operator
```


### Filter function

Filter data of countries from 1997 and with Oceania continent
```{r}
filter(countries, year==1997, continent=="Oceania")
```


Filter data of countries from 2007 and with gdpPercap greater than $40,000
```{r}
filter(countries, year==2007, gdpPercap>40000)
```


Question: What relational operator should you use for greater than and equal to $40,000?
```{r}
filter(countries, year==2007, gdpPercap>=40000)
```


Filter observations from year 2002 or year 2007
```{r}
filter(countries, year==2002 | year==2007) # Use vertical line for Or condition
```



### Arrange function

Arrange countries based on life expectancy (from smallest to largest)
```{r}
arrange(countries, lifeExp)
```


Arrange countries based on life expectancy (from smallest to largest), sorted first by year (from earliest year available to most recent)
```{r}
arrange(countries, year, lifeExp)
```


Arrange countries based on life expectancy (from largest to smallest)
```{r}
arrange(countries, desc(lifeExp))
```


Arrange countries based on life expectancy (from largest to smallest), sorted first by year (from earliest year available to most recent)
```{r}
arrange(countries, year, desc(lifeExp))
```


### Summarise function

Available sub-functions:

* Center: mean(), median()
* Spread: sd(), IQR(), mad()
* Range: min(), max(), quantile()
* Position: first(), last(), nth(),
* Count: n(), n_distinct()
* Logical: any(), all()


Find average value of the lifeExp variable
```{r}
summarise(countries, mean_lifeExp=mean(lifeExp))
```


Find range of the lifeExp variable
```{r}
summarise(countries, range_lifeExp=max(lifeExp)-min(lifeExp))
```


### Mutate function

Create a new variable called 'GDP' (a product of 'gdpPercap' and 'pop')
```{r}
mutate(countries, GDP=gdpPercap*pop)
```




## Piping with Multiple Functions

What if you want to perform multiple functions in R? Use Pipe operator (%>%) in the dplyr package. It allows you to perform multiple functions without using nested parentheses. 


This is how piping looks like:
```{r}
# DataFrame%>%
    #function to execute first %>%
        #function to execute second %>%
            #function to execute third
```


Example of Piping:

Finding average life expectancy in each continent in 2007

* Select continent, year, lifeExp
* Filter observations from 2007
* Split data by continents
* Summarize the mean life expectancy within each continent
* Round the average life expectancy to 2 decimal points


```{r}
countries %>%
  select(continent, year, lifeExp) %>% #select continent,  year, lifeExp
    filter(year==2007) %>% #filter observations from 2007
      group_by(continent) %>% # split data by continents
          summarise(mean_lifeExp=mean(lifeExp)) %>% #summarize the mean life expectancy within each continent
            mutate(mean_lifeExp=round(mean_lifeExp,2)) # Use mutate to modify existing variable, round mean weight to 2 decimal points
```


### Exercise #2:

Finding the average of GDP per Capita of Europen countries

* Filter observations of European countries
* Group by year
* Summarize the mean gdpPercap


Answers:
```{r}
countries %>%
  filter(continent=="Europe") %>%
      group_by(year) %>%
      summarise(mean_gdpPercap=mean(gdpPercap))
```



# Data visualization with ggplot


## Barchart

Shows the distribution of a categorical variable
```{r}
ggplot(data=countries, aes(x=continent, fill=continent))+
    geom_bar()+
    labs(title="Counts of observations by continents", x="Continents", y="Number of observations")
```


Modify count to measure the number of countries by dividing by 12
```{r}
ggplot(data=countries, aes(x=continent, fill=continent))+
    geom_bar(aes(y=..count../12))+
    labs(title="Data distribution by continents", x="Continents", y="Number of countries")
```




## Histogram

Visualizes the distribution of one quantitative variable
```{r}
ggplot(data=countries, aes(x=lifeExp)) + 
    geom_histogram(binwidth=4,color="black", fill="lightblue") +
    labs(title="Histogram of life expenctancy ", x="Life expectancy", y="Number of observations")
```



## Boxplots

Boxplots are used to compare distributions of one quantitative variable across multiple categories (a visualization alternative to density plots)

Comparing distribution of lifeExp variable by continent:
```{r}
ggplot(data=countries, aes(x=continent, y=lifeExp, fill=continent))+
    geom_boxplot(outlier.size=1)+
    labs(title="Boxplots of life expenctancy by continent")
```





## Scatterplots

Scatterplots show the relationship between two quantitative variables
```{r}
ggplot(data=countries, aes(x=gdpPercap, y=lifeExp))+ 
    geom_point()+
    labs(title="Scatterplot of gdpPercap and lifeExp")
```


Add color to represent different continents
```{r}
ggplot(data=countries, aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent))+
    labs(title="Scatterplot of gdpPercap and lifeExp")
```


Transform the gdpPercap variable using a log10 scale
```{r}
ggplot(data=countries, aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent))+
    scale_x_log10()+
    labs(title="Scatterplot of gdpPercap(log10) and lifeExp", x="Log 10 of gdpPercap")
```


Add a best-fit line to the scatterplot
```{r}
ggplot(data=countries, aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent))+
    geom_smooth(method="lm")+
    scale_x_log10()+
    labs(title="Scatterplot of gdpPercap(log10) and lifeExp", x="Log 10 of gdpPercap")
```



## Piping in ggplot

Make a timeplot of `lifeExp` for European countries with a population above 15m
```{r}
countries %>%
    filter(continent=="Europe", pop>15000000) %>% 
    ggplot(aes(x=year, y=lifeExp, color=country))+
    geom_line(size=1)+ 
    geom_point()+  
    labs(title="Timeplot of average life expectancy by continent")
```


### Exercise #3

Make a timeplot of `gdpPercap` for countries in Americas with population above 20m

Hints:

* Use filter function to limit the selection of data to Americas, pop > 20000000
* Use ggplot, geom_point, and geom_line to make a graph


Visualize historical changes in average life expectancy of European countries with a population above 15m
```{r}
gdp_timeplot <- countries %>%
    filter(continent=="Americas", pop>20000000) %>% 
    ggplot(aes(x=year, y=gdpPercap, color=country))+
    geom_line(size=1)+ 
    geom_point()+  
    labs(title="Timeplot of GDP per capita of selected countries in the Americas")

gdp_timeplot
```


## Save produced graphs

```{r}
ggsave("gdp_timeplot.png", plot = gdp_timeplot, device="png", width = 8, height = 5)
```



## Save the dataframe as a CSV file 

Save the dataframe as a csv file
```{r}
write.csv(countries, file="countries.csv")
```
